---
title: "Tidyverse"
subtitle: "Data cleaning and wrangling in R"
output: 
  html_document:
    toc: TRUE
    df_print: paged
    number_sections: FALSE
    highlight: tango
    theme: lumen
    toc_depth: 3
    toc_float: true
    css: custom.css 
    self_contained: false
    includes:
      after_body: footer.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

------------------------------------------------------------------------

# Data manipulation with `dplyr`

A second fundamental package of the tidyverse is called `dplyr`. In this section you'll learn and practice examples using some functions in `dplyr` to work with data. Those are:

-   `select()`: keep or exclude some columns
-   `filter()`: keep rows that satisfy your conditions
-   `mutate()`: add columns from existing data or edit existing columns
-   `group_by()`: lets you define groups within your data set
-   `summarize()`: get summary statistics
-   `arrange()`: reorders the rows according to single or multiple variables

```{r, echo=FALSE}
library(palmerpenguins)
library(tidyverse)
```

To demonstrate and practice how these verbs (functions) work, we'll use the penguins dataset.

The 3 species of penguins in this data set are Adelie, Chinstrap and Gentoo. The data set contains 8 variables:

-   **species:** a factor denoting the penguin species (Adelie, Chinstrap, or Gentoo)
-   **island:** a factor denoting the island (in Palmer Archipelago, Antarctica) where observed
-   **culmen_length_mm:** a number denoting length of the dorsal ridge of penguin bill (millimeters)
-   **culmen_depth_mm:** a number denoting the depth of the penguin bill (millimeters)
-   **flipper_length_mm:** an integer denoting penguin flipper length (millimeters)
-   **body_mass_g:** an integer denoting penguin body mass (grams)
-   **sex:** a factor denoting penguin sex (MALE, FEMALE)
-   **year** an integer denoting the year of the record

## `select()`

The first verb (function) we will utilize is `select()`. We can employ it to manipulate our data based on **columns**. If you recall from our initial exploration of the data set there were eight variables attached to every observation. Do you recall them? If you do not, there is no problem. You can utilize `names()` to retrieve the names of the variables in a data frame.

```{r}
names(penguins)
```

Say we are only interested in the species, island, and year variables of these data, we can utilize the following syntax:

`select(data, columns)`

**Excercise 3** *The following code chunk would select the variables we need. Can you adapt it, so that we keep the body_mass_g and sex variables as well?*

```{r,, eval=F}
dplyr::select(penguins, species, island, year)
```

Good to know: To **drop** variables, use `-` before the variable name, i.e. select(penguins, -year) to drop the year column (select everything but the year column).

## `filter()` `r emo::ji("coffee")`

The second verb (function) we will employ is `filter()`. `filter()` lets you use a logical test to extract specific **rows** from a data frame. To use `filter()`, pass it the data frame followed by one or more logical tests. `filter()` will return every row that passes each logical test.

The more commonly used logical operators are:

-   `==`: Equal to
-   `!=`: Not equal to
-   `>`, `>=`: Greater than, greater than or equal to
-   `<`, `<=`: Less than, less than or equal to
-   `&`, `|`: And, or

Say we are interested in retrieving the observations from the year 2007. We would do:

```{r , eval=F}
dplyr::filter(penguins, year == 2007)
```

**Excercise 4** Can you adapt the code to retrieve all the observations of Chinstrap penguins from 2007

```{r, echo=FALSE, eval=F}
# you just need to utilize & and type the logical operator for the species
dplyr::filter(penguins, year == 2007 & species == "Chinstrap")
```

**Excercise 5** We can leverage the pipe operator to sequence our code in a logical manner. Can you adapt the following code chunck with the pipe and conditional logical operators we discussed?

```         
only_2009 <- dplyr::filter(penguins, year == 2009)
only_2009_chinstraps <- dplyr::filter(only_2009, species == "Chinstrap")
only_2009_chinstraps_species_sex_year <- dplyr::select(only_2009_chinstraps, species, sex, year)
final_df <- only_2009_chinstraps_species_sex_year
final_df #to print it in our console
```

```{r, echo=FALSE, eval=F}
penguins |>
  dplyr::filter(year == 2009 & species == "Chinstrap") |>
  dplyr::select(species, sex, year)
```

## `mutate()` ``` r emo::ji("closed_umbrella")``r emo::ji("open_umbrella") ```

`mutate()` lets us create, modify, and delete columns. The most common use for now will be to create new variables based on existing ones. Say we are working with a U.S. American client and they feel more confortable with assessing the weight of the penguins in pounds. We would utilize `mutate()` as such:

`mutate(new_var_name = manipulated old_var(s))`

```{r,, eval=F}
penguins |>
  dplyr::mutate(body_mass_lbs = body_mass_g/453.6)
```

## `group_by()` and `summarize()`

These two verbs `group_by()` and `summarize()` tend to go together. When combined , 'summarize()\` will create a new data frame. It will have one (or more) rows for each combination of grouping variables; if there are no grouping variables, the output will have a single row summarising all observations in the input. For example:

```{r}
# compare this output with the one below
penguins |>
  dplyr::summarize(heaviest_penguin = max(body_mass_g, na.rm = T)) 
```

```{r}
penguins |>
  dplyr::group_by(species) |>
  dplyr::summarize(heaviest_penguin = max(body_mass_g, na.rm = T))
```

There is also an alternate approach to calculating grouped summary statistics called per-operation grouping. This allows you to define groups in a .by argument, passing them directly in the `summarise()` call. These groups don't persist in the output whereas the ones used with `group_by` do. You can read more about both these approaches in [R for Data Science, 2nd edition](https://r4ds.hadley.nz).

```{r}
penguins |>
  dplyr::summarise(heaviest_penguin = max(body_mass_g, na.rm = T), .by = species)

penguins |>
  dplyr::summarise(heaviest_penguin = max(body_mass_g, na.rm = T), .by = c(species, sex))
```

**Excercise 6** Can you get the weight of the lightest penguin of each species? You can use `min()`. What happens when in addition to species you also group by year `group_by(species, year)`?

```{r, eval = F}
penguins |>
  dplyr::group_by(species, year) |>
  dplyr::summarize(lightest_penguin = min(body_mass_g, na.rm = T))
```

## `arrange()` ``` r emo::ji("egg")``r emo::ji("hatching_chick")``r emo::ji("hatched_chick") ```

The `arrange()` verb is pretty self-explanatory. `arrange()` orders the rows of a data frame by the values of selected columns in ascending order. You can use the `desc()` argument inside to arrange in descending order. The following chunk arranges the data frame based on the length of the penguins' bill. You hint tab contains the code for the descending order alternative.

```{r, eval=F}
penguins |>
  dplyr::arrange(bill_length_mm)

```

```{r, eval=F}
penguins |>
  dplyr::arrange(desc(bill_length_mm))

```

**Excercise 6** Can you create a data frame arranged by body_mass_g of the penguins observed in the "Dream" island?

```{r, echo = FALSE, eval=F}
penguins |>
  dplyr::filter(island == "Dream") |>
  dplyr::arrange(body_mass_g)
```

**Quiz**

1.  Which verb allows you to index columns?

-   select()
-   filter()
-   summarize()
-   group_by()

2.  Which verb allows you to index rows?

-   select()
-   filter()
-   summarize()
-   group_by()

3.  How long was the longest observed bill of a Gentoo penguin in 2008?

4.  How long was the shortest observed bill of a Gentoo penguin in 2008?

## optional: Other dplyr functions

`ungroup()`: For ungrouping data after using the `group_by()` command - Particularly useful with the `summarise` and `mutate` commands, as we've already seen.

`slice()`: Subset rows by position rather than filtering by values. - E.g. `penguins |> slice(c(1, 5))`

`pull()`: Extract a column from as a data frame as a vector or scalar. - E.g. `penguins |> filter(gender=="female") |> pull(height)`

`count()` and `distinct()`: Number and isolate unique observations. - E.g. `penguins |> count(species)`, or `penguins |> distinct(species)` - You could also use a combination of `mutate`, `group_by`, and `n()`, e.g. `penguins |> group_by(species) |> mutate(num = n())`.

The final set of dplyr verbs we'd like you to know are the family of join operations. These are important enough that we want to go over some concepts in a bit more depth.

However - note that we will cover relational data structures (and SQL) in even more depth in a separate lab session!

------------------------------------------------------------------------

# Joins with `dplyr` (optional)

One of the mainstays of the dplyr package is merging data with the family [join operations](https://cran.r-project.org/web/packages/dplyr/vignettes/two-table.html).

-   `inner_join(df1, df2)`
-   `left_join(df1, df2)`
-   `right_join(df1, df2)`
-   `full_join(df1, df2)`
-   `semi_join(df1, df2)`
-   `anti_join(df1, df2)`

(You might find it helpful to to see visual depictions of the different join operations [here](https://r4ds.had.co.nz/relational-data.html).)

For the simple examples that I'm going to show here, we'll need some data sets that come bundled with the [**nycflights13**](http://github.com/hadley/nycflights13) package. - Load it now and then inspect these data frames in your own console.

```{r, echo = F}
library(nycflights13)
#flights 
#planes
```

Let's perform a [left join](https://stat545.com/bit001_dplyr-cheatsheet.html#left_joinsuperheroes-publishers) on the flights and planes datasets. - *Note*: I'm going subset columns after the join, but only to keep text on the slide.

```{r}
left_join(flights, planes) |>
  select(year, month, day, dep_time, arr_time, carrier, flight, tailnum, type, model)|>
  head(3) ## Just to save vertical space in output
```

Note that dplyr made a reasonable guess about which columns to join on (i.e. columns that share the same name). It also told us its choices:

```         
## Joining, by = c("year", "tailnum")
```

However, there's an obvious problem here: the variable "year" does not have a consistent meaning across our joining datasets! - In one it refers to the *year of flight*, in the other it refers to *year of construction*.

Luckily, there's an easy way to avoid this problem. - Try `?dplyr::join`.

You just need to be more explicit in your join call by using the `by =` argument. - You can also rename any ambiguous columns to avoid confusion.

```{r}
left_join(
  flights,
  planes |> rename(year_built = year), ## Not necessary w/ below line, but helpful
  by = "tailnum" ## Be specific about the joining column
  ) |>
  select(year, month, day, dep_time, arr_time, carrier, flight, tailnum, year_built, type, model) |>
  head(3) 
```

Note what happens if we again specify the join column... but don't rename the ambiguous "year" column in at least one of the given data frames.

```{r join3}
left_join(
  flights,
  planes, ## Not renaming "year" to "year_built" this time
  by = "tailnum"
  ) |>
  select(contains("year"), month, day, dep_time, arr_time, carrier, flight, tailnum, type, model) |>
  head(3)
```

Make sure you know what "year.x" and "year.y" are. Again, it pays to be specific.

Now let's take another example using the palmerpenguins dataset we used earlier today. Suppose we have the following additional information on the three islands the penguins are from:

```{r}
islands <- data.frame(name = c("Torgersen", "Biscoe", "Dream"), 
                      coordinates = c("64°46′S 64°5′W","65°26′S 65°30′W","64°44′S 64°14′W"))
```

We want to merge these datasets using the column for island names, however the columns are named differently in both datasets. This is where dplyr's new updates come in handy.

dpyr 1.1.0 has introduced a whole [new bunch of extensions](https://www.tidyverse.org/blog/2023/01/dplyr-1-1-0-joins/) in the `*_join` family of functions that bring their functionality closer to data.table, SQL etc. One of them is the `join_by` function for the `by` argument.

Previously, we'd merge these data frames as follows:

```{r}
left_join(
  penguins,
  islands,
  by = c("island" = "name")
  ) |>
  head(3)
```

With the `join_by` function, we would achieve the same results as follows:

```{r}
left_join(
  penguins,
  islands,
  by = join_by(island == name)
  ) |>
  head(3)

#or simply
left_join(
  penguins,
  islands,
  join_by(island == name)
  ) |>
  head(3)
```

This approach is recommended for its succinctness and ease of reading, similar to logical statements.

You can check out additional functionalities of the different `*_join()` functions such as new arguments for handling `multiple` matches and `unmatched` rows, as well as for defining the `relationship` between data frames in [R for Data Science, 2nd Edition] (<https://r4ds.hadley.nz/joins>).

# Actually learning R `r emo::ji("backpack")`

Let us remind you again, the key to learning `R` is: **Google**! We can only give you an overview over basic `R` functions, but to really learn `R` you will have to actively use it yourself, trouble shoot, ask questions, and google! It is very likely that someone else has had the exact same or just *similar enough* issue before and that the R community has answered it with 5+ different solutions years ago. `r emo::ji("wink")`

------------------------------------------------------------------------

# Coding style `r emo::ji("art")`

## Why adhere to a particular style of coding?

-   It reduces the number of arbitrary decisions you have to consciously make during coding. We make an arbitrary decision (convention) once, not always ad hoc.
-   It provides consistency.
-   It makes code easier to write.
-   It makes code easier to read, especially in the long term (i.e. two days after you've closed a script).

## What are questions of style?

-   Questions of style are a matter of opinion.
-   We will mostly follow Hadley Wickham's opinion as expressed in the "[tidyverse style guide](https://style.tidyverse.org/)".
-   We'll consider how to
    -   name,
    -   comment,
    -   structure, and
    -   write.

## Naming things

### Naming files

-   Code file names should be meaningful and end in `.R`.
-   Avoid using special characters in file names. Stick with numbers, letters, dashes (`-`), and underscores (`_`).
-   Some examples:

``` bash
# Good
fit_models.R
utility_functions.R

# Bad
fit models.R
foo.r
stuff.r
```

-   If files should be run in a particular order, prefix them with numbers:

``` bash
00_download.R
01_explore.R
...
09_model.R
10_visualize.R
```

### Naming objects and variables

-   There are various conventions of how to write phrases without spaces or punctuation. Some of these have been adapted in programming, such as [camelCase](https://en.wikipedia.org/wiki/Camel_case), [PascalCase](https://techterms.com/definition/pascalcase), or [snake_case](https://en.wikipedia.org/wiki/Snake_case).
-   The [`tidyverse`](https://style.tidyverse.org/syntax.html#object-names) way: Object and variable names should use only lowercase letters, numbers, and underscores.
-   Examples:

``` bash
# Good
day_one # snake_case
day_1 # snake_case

# Less good
dayOne # camelCase
DayOne # PascalCase
day.one # dot.case

# Dysfunctional
day-one # kebab-case
```

### Naming functions

-   In addition to following the general advice for object names, strive to use verbs for function names:

``` bash
# Good
add_row()
permute()

# Bad
row_adder()
permutation()
```

-   Also, try avoiding function names that already exist, in particular those that come with a loaded package.
-   This often implies a trade-off between shortness and uniqueness. In any case, you would try to avoid situations that force you disambiguate functions with the same name (as in `dplyr::select`; see ["R packages"](https://r-pkgs.org/dependencies-mindset-background.html#sec-dependencies-namespace)).
-   Check out this [Wikipedia page](https://en.wikipedia.org/wiki/Naming_convention_(programming)) or this [Stackoverflow post](https://stackoverflow.com/questions/17326185/what-are-the-different-kinds-of-cases) for more background on naming conventions in programming!
-   For more good advice on how to name stuff, see [this legendary presentation](http://www2.stat.duke.edu/~rcs46/lectures_2015/01-markdown-git/slides/naming-slides/naming-slides.pdf) by Jenny Bryan.

## Commenting on things `r emo::ji("comment")`

### Why comment on things at all?

-   It's often tempting to set up a project assuming that you will be the only person working on it, e.g. as homework. But that's almost never true.
-   You have project partners, co-authors, principals.
-   Even if not, there's someone else who you always have to keep happy: Future-you.
-   Comment often to make Future-you happy about Past-you by document what Present-You is doing/thinking/planning to do.

### General advice

-   Each line of a comment should begin with the comment symbol and a single space: `#`
-   Use comments to record important findings and analysis decisions.
-   If you need comments to explain what your code is doing, consider rewriting your code to be clearer.
-   But: comments can work well as "sub-headlines".
-   If you discover that you have more comments than code, consider switching to R Markdown.
-   (Longer) comments generally work better if they get their own line.

```{r, eval = FALSE}
# define job status
dat$at_work <- dat$job %in% c(2, 3)
dat$at_work <- dat$job %in% c(2, 3) # define job status
```

```{r, eval = FALSE}
penguins |>
  filter(species == "Gentoo" | species == "Adelie")

penguins |> 
  filter(species %in% c("Gentoo", "Adelie"))
```

### Giving structure

-   Use commented lines together with dashes to break up your file into easily readable chunks.
-   RStudio automatically detects these chunks and turns them into sections in the script outline.

```{r, eval = FALSE}
# Input/output ---------------------

# input
c("data/survey2021.csv")

#  output
c("survey_2021_cleaned.RData",
  "resp_ids.csv")

# Load data ------------------------

# Plot data ------------------------
```

## Other stuff `r emo::ji("package")`

-   Use **spaces** generously, but not too generously. Always put a space after a comma, never before, just like in regular English.
-   Use `<-`, not `=`, for **assignment**.
-   For **logical operators**, prefer `TRUE` and `FALSE` over `T` and `F`.
-   To facilitate readability, **keep your lines short**. Strive to limit your code to about 80 characters per line.
-   If a **function call is too long** to fit on a single line, use one line each for the function name, each argument, and the closing bracket.
-   Use **pipes**. When you use them, they should always have a space before it, and should usually be followed by a new line.

**Spacing**

```{r, eval = FALSE}
# Good
mean(x, na.rm = TRUE)
height <- (feet * 12) + inches

# Bad
mean(x,na.rm=TRUE) 
mean ( x, na.rm = TRUE )
height<-feet*12+inches
```

**Piping**

```{r, eval = FALSE}
babynames |>
  filter(name |> equals("Kim")) |>
  group_by(year, sex) |>
  summarize(total = sum(n)) |>
  qplot(year, total, color = sex, data = ., 
        geom = "line") |>
  add(ggtitle('People named "Kim"')) |>
  print
```

------------------------------------------------------------------------

# Sources {.unnumbered}

This tutorial is partly based on [*R for Data Science*](https://r4ds.hadley.nz/), section 5.2, [*Quantitative Politics with R*](http://qpolr.com/data.html/), chapter 3, the [Tidyverse Session](https://github.com/uo-ec607/lectures/tree/master/05-tidyverse) in the course Data Science for Economists by Grant McDermott, and [*Teaching the Tidyverse in 2023*](https://www.tidyverse.org/blog/2023/08/teach-tidyverse-23/).

------------------------------------------------------------------------

# Appendix

## Other dplyr goodies `r emo::ji("gift")`

### `slice()`

Subset rows by position rather than filtering by values. There's also `slice_sample()` to randomly select rows, `slice_head()` and `slice_tail()`to select first or last rows, and more.

```{r}
starwars |> slice(c(1, 5))
```

### `pull()`

Extract a column from as a data frame as a vector or scalar.

```{r}
starwars |> filter(gender=="feminine") |> pull(height)
```

### `count()`, `n()`, `distinct()`

Count, number and isolate unique observations.

```{r}
starwars |> count(species) |> head(6)
starwars |> distinct(species) |> pull() |> sort() |> magrittr::extract(1:5) 
```

You could also use a combination of `mutate()`, `group_by()`, and `n()`, e.g. `starwars |> group_by(species) |> mutate(num = n())`.

### `where()`

Select the variables for which a function returns true.

```{r}
starwars |> select(where(is.numeric)) |> names()
```

### `across()`

Summarize or mutate multiple variables in the same way. More information [here](https://dplyr.tidyverse.org/reference/across.html).

```{r}
starwars |> 
  mutate(across(where(is.numeric), scale)) |> 
  head(3)
```

### `case_when()`

Vectorize multiple `if_else()` (or base R `ifelse()`) statements.

```{r}
starwars |> 
  mutate( 
    height_cat = case_when(
      height < 160 ~ "tiny",
      height >= 160 & height < 190 ~ "medium",
      height >= 190 & height < 220 ~ "tall",
      height >= 220 ~ "giant"
    )
  ) |>
  pull(height_cat) |> table()
```

### Window functions

There are also a whole class of [window functions](https://cran.r-project.org/web/packages/dplyr/vignettes/window-functions.html) for getting leads and lags, ranking, creating cumulative aggregates, etc. See `vignette("window-functions")`.

## Other tidyr goodies `r emo::ji("broom")`

### `crossing()`

Get the full combination of a group of variables.

```{r cross1}
crossing(side = c("left", "right"), height = c("top", "bottom"))
```

### `expand()` and `complete()`

See `?expand()` and `?complete()` for more specialized functions that allow you to fill in (implicit) missing data or variable combinations in existing data frames. Base R alternative: `expand.grid()`.

### `drop_na(data, ...)`

Drop rows containing NAs in `...` columns.

### `fill(data, ..., direction = c("down", "up"))`

Fill in NAs in `...` columns with most recent non-NA values.

## Pipes `r emo::ji("sunglasses")`

### The beauty of pipes

-   The forward-pipe operator `|>` pipes the left-hand side values forward into expressions on the right-hand side.
-   It serves the natural way of reading ("do this, then this, then this, ...").
-   We replace `f(x)` with `x |> f()`.
-   It avoids nested function calls.
-   It minimizes the need for local variables and function definitions.

**The pipe way**

```{r, eval = FALSE}
Alex |>
  wake_up(7) |>
  shower(temp = 38) |>
  breakfast(c("coffee", "croissant")) |>
  walk(step_function()) |>
  bvg(
    train = "U2",
    destination = "Stadtmitte"
  ) |>
  hertie(course = "Intro to DS")
```

**The classic way**

```{r, eval = FALSE}
hertie(
  bvg(
    walk(
      breakfast(
        shower(
          wake_up(
            Alex, 7
          ),
          temp = 38
        ),
        c("coffee", "croissant")
      ),
      step_function()
    ),
    train = "U2",
    destination = "Stadtmitte"
  ),
  course = "Intro to DS"
)
```

**The classic way, nightmare edition**

```{r, eval = FALSE}
alex_awake <- wake_up(Alex, 7)
alex_showered <- shower(alex_awake, 
                        temp = 38)
alex_replete <- breakfast(alex_showered, 
                          c("coffee", "croissant"))
alex_underway <- walk(alex_replete, 
                      step_function())
alex_on_train <- bvg(alex_underway, 
                         train = "U2", 
                         destination = "Stadtmitte")
alex_hertie <- hertie(alex_on_train, 
                      course = "Intro to DS")
```

### Piping etiquette

-   Pipes are not very handy when you need to manipulate more than one object at a time. Reserve pipes for a sequence of steps applied to one primary object.
-   Don't use the pipe when there are meaningful intermediate objects that can be given informative names (and that are used later on).
-   `|>` should always have a space before it, and should usually be followed by a new line.

### The base R pipe: \|\>

The magrittr pipe has proven so successful and popular that the R core team [recently added](https://stat.ethz.ch/R-manual/R-devel/library/base/html/pipeOp.html) a "native" pipe operator to base R (version 4.1), denoted `|>`. Here's how it works:

``` r
mtcars |> subset(cyl == 4) |> head()
mtcars |> subset(cyl == 4) |> (\(x) lm(mpg ~ disp, data = x))()
```

Now, should we use the "magrittr" pipe or the native pipe? The native pipe might make more sense in the long term, since it avoids dependencies and might be more efficient. Check out [this Stackoverflow post](https://stackoverflow.com/questions/67633022/what-are-the-differences-between-rs-new-native-pipe-and-the-magrittr-pipe) for a discussion of differences.
